Accelerating adaptive ultrasound algorithms using graphics processing units \vspace{0.5cm}

### 2014-03-30 ThesisJP/Background/background.tex
Added several images. Text has been updated.

\section {Medical ultrasound imaging}\label{sec:ultrasound}
Medical ultrasound imaging is considered to be a low-risk tool for non-invasive investigation. Neither gamma rays nor high intensity magnetic fields are involved, just sound pressure below well-controlled limits. It is also far less costly than other medical imaging modalities and offers real-time inspection of the human body.  Ultrasound imaging is well known for its use in fetal imaging during pregnancy, however, in this thesis we will focus on a different modality, namely cardiac ultrasound imaging. An illustration of the human heart is therefore included in Fig.\,\ref{fig:human_heart}. Throughout the thesis we will refer to some of the parts shown in Fig.\,\ref{fig:human_heart}. Especially note the four chambers of the heart, the left and right atrium, and the left and right ventricle. The tissue separating the left and right ventricle is know as the inter-ventricular septum. The left side is responsible of pumping blood out to the whole body through the aortic valve. It is therefore the strongest and hardest working part of the heart. When blood flows out of the aorta it is important that the mitral valve, which separates the left atrium from the left ventricle, is closed and does not leak. The left side, and especially the left ventricle and the mitral valve, therefore receives the most attention during cardiac ultrasound examinations. When visualizing 3D ultrasound images in \textbf{Paper IV} we will have a strong focus on visibility of cardiac tissue, thus note for later that the endocardium refers to the innermost layer of the cardiac muscle.

\begin{figure}
\centering
\subfigure[]{
	\includegraphics[width=0.47\textwidth]{img/Diagram_of_the_human_heart.png}\label{fig:human_heart}
}
\subfigure[]{
	\includegraphics[width=0.47\textwidth]{img/ultrasound_grid.eps}\label{fig:us_grid}
}
\caption{a) Overview of the human heart. Illustration from wikipedia.org. b) Coordinate system for phased array imaging.}
\end{figure}

@@

To generate the ultrasonic frequencies used for imaging a piezoelectric ceramic is used. A piezoelectric material is special in that way that ultrasound vibrations are generated when voltage is applied and \textit{vice versa}. It can therefore be used to both send and receive ultrasound signals, like a combined microphone and speaker. Piezoelectric elements organized in an array, is referred to as an ultrasound \textit{probe}, and a special probe is typically designed for each medical imaging modality. %For cardiac ultrasound imaging the probe must not only be designed for a certain frequency range, its footprint also needs to be smaller than the intercostal window in order to avoid shadowing by ribs. 
A line in an ultrasound image is generated by firing a ultrasound pulse into the body and sample the received echo. This amplitude trace is known as an A-mode image.

\subsection{Beamforming}
If the width of a piezoelectric element is less than the wavelength of the transmitted signal the element is said to be \textit{omnidirectional}, hence energy is distributed or sensed equally in all directions. When multiple omnidirectional elements are arranged in a linear array the transmitted pulse will get shaped into a beam\footnote{A beam refers to the pulse in space integrated across time.} of ultrasound by diffraction as the size of the array  (the \textit{aperture}) grows. The 3 dB angular width of this beam, and also the angular resolution of the array, is given by \cite{AngelUltrasound}:
\begin{align}\label{eq:res}
\delta\theta_a = \frac{\lambda}{D} = \frac{c}{fD},
\end{align}
\nomi{$\delta\theta_a$}{Angular resolution of an aperture}where \nom{$\lambda$}{Wavelength} and \nom{$f$}{Frequency} is the ultrasound wavelength and frequency respectively, $c$ is the speed of sound, and \nom{$D$}{Aperture size} is the aperture width. With an average speed-of-sound in human tissue of $1540\frac{\text{m}}{\text{s}}$, the angular resolution or opening angle of a 2 cm cardiac probe operating at 3 MHz is around 1.5\degree, which at 8 cm range gives a lateral resolution of around 2 mm. The axial resolution is inverse proportional to the pulse length, and is typically better than the lateral resolution.

@@

If time delays are applied on the signal sent to the different elements it is possible to steer and focus the ultrasound beam in space (Fig.\,\ref{fig:das}). This technique is known as \textit{beamforming}, and for ultrasound imaging it is traditionally applied both when transmitting and receiving ultrasound pulses. For medical ultrasound imaging there are two main types of ultrasound probes, linear and phased probes \cite{AngelUltrasound}. A cardiac probe is a phased probe, which means that the image is constructed from beams, or A-mode images, which are gradually steered from $-\theta/2$ to $\theta/2$ radians, resulting in an imaging sector which are $\theta$ radians wide. This kind of image, where A-mode images are acquired from multiple spatial locations in the $xz$-plane, is known as a B-mode image (B for brightness). If A-mode images are acquired from the $yz$-plane as well we get what is known as a volumetric B-mode or 3D ultrasound image. A probe for 3D imaging has several thousand elements arranged as a matrix. In this thesis we will focus on linear arrays for 2D imaging when the topic is adaptive beamforming, and 3D imaging when the topic is adaptive visualization.

If \nom{$M$}{Number of array elements} is the number of elements in the array, $x_m$ is the per-element data, and \nom{$\Delta_m$}{Per-element focusing and steering delays} is the per-element focusing and steering delay, the standard form of beamforming used on receive in ultrasound imaging is:
\begin{align}\label{eq:das}
z[n] = \sum_{m = 0}^{M-1}w_m^*x_m[n - \Delta_m[n]] = \vec{w}^H\vec{x}[n].
\end{align}
This is known as \textit{delay-and-sum} (DAS) beamforming. Here, \nom{$\vec{w}$}{Array weight vector (apodization)} is the array \textit{weight vector} or \textit{apodization}, $\vec{x}$ is a pre-delayed vector of element data, and $z[n]$ is the output or the A-mode image in the direction given by $\Delta_m[n]$.  From now on $x_m[n]$ will refer to its delayed version $x_m[n-\Delta_m[n]]$. A schematic overview of delay-and-sum array beamforming is depicted in Fig.\,\ref{fig:das}. If $p_o$ is the phase center of our array and $p$ is the point where we want to steer and focus the ultrasound beam, we have to use the following delays:
\begin{align}
\Delta_m[n] =\frac{|p_o+p| - |p_m+p|}{c},
\end{align}
where \nom{$c$}{Speed of sound} is the speed of sound in human tissue, and $p_m$ is the position of the m'th element. On transmit the inverse set of delays has to be used. Also note that on transmit we have to decide where to steer and focus prior to transmitting a pulse, transmit delays therefore do not vary within an A-mode image.

\subsection{Sampling and processing complexity}
When generating a B-mode image we have to insure that the spatial and temporal sampling are sufficient. We have already seen that the lateral resolution of an ultrasound probe is given by (\ref{eq:res}).  Hergum \textit{et al.} \cite{Hergum2009} have further shown that the lateral frequency support for a system which both emit and receive energy is governed by the sum of the transmit (\nom{tx}{Transmit}) and receive (\nom{rx}{Receive}) aperture. The lateral resolution of what is known as an active imaging system is found to be
\begin{align}
\delta\theta = \frac{\lambda}{D_{tx} + D_{rx}}.
\end{align}
\nomi{$\delta\theta$}{Angular resolution of a two-way imaging system}\nomi{$D_{tx}$}{Aperture size on transmit}\nomi{$D_{rx}$}{Aperture size on receive}Then, in order to properly sample an imaging sector of $\theta$ radians we have to acquire 
\begin{align}
N_\theta = \frac{\theta}{\delta\theta}
\end{align}
\nomi{$N_\theta$}{Number of lines required for proper sampling of an imaging sector}A-mode images uniformly distributed across the sector. If \nom{$r$}{Imaging range.} is the maximum range we want to image we get an imaging frame rate of
\begin{align}
f_{\text{FPS}} = \frac{c}{2rN_\theta}
\end{align} 
frames per second (\nom{FPS}{Frames per second}). With our 2 cm cardiac probe we get a frame rate of around $55$ FPS when imaging a 70\degree{} sector with $r=15$ cm.

>>>>

\section {Medical ultrasound imaging}\label{sec:ultrasound}
Medical ultrasound imaging is considered to be a low-risk tool for non-invasive investigation. Neither gamma rays nor high intensity magnetic fields are involved, just sound pressure below well-controlled limits. It is also far less costly than other medical imaging modalities and offers real-time inspection of the human body.  Ultrasound imaging is well known for its use in fetal imaging during pregnancy, however, in this thesis we will focus on a different modality, namely cardiac ultrasound imaging. An illustration of the human heart is therefore included in Fig.\,\ref{fig:human_heart}. Throughout the thesis we will refer to some of the parts shown in Fig.\,\ref{fig:human_heart}. Especially note the four chambers of the heart, the left and right atrium, and the left and right ventricle. The tissue separating the left and right ventricle is known as the inter-ventricular septum. The left side is responsible of pumping blood out to the whole body through the aortic valve. It is therefore the strongest and hardest working part of the heart. When blood flows out of the aorta it is important that the mitral valve, which separates the left atrium from the left ventricle, is closed and does not leak. The left side, and especially the left ventricle and the mitral valve, therefore receives the most attention during cardiac ultrasound examinations. When visualizing 3D ultrasound images in \textbf{Paper IV} we will have a strong focus on visibility of cardiac tissue, thus note for later that the endocardium refers to the innermost layer of the cardiac muscle.

\begin{figure}
\centering
\subfigure[]{
	\includegraphics[width=0.47\textwidth]{img/Diagram_of_the_human_heart.png}\label{fig:human_heart}
}
\subfigure[]{
	\includegraphics[width=0.47\textwidth]{img/ultrasound_grid.pdf}\label{fig:us_grid}
}
\caption{a) Overview of the human heart. Illustration from wikipedia.org. b) Illustration of probe geometry and layout for phased array imaging.}
\end{figure}

@@

To generate the ultrasonic frequencies used for imaging, a piezoelectric ceramic is used. A piezoelectric material is special in that way that ultrasound vibrations are generated when voltage is applied and \textit{vice versa}. It can therefore be used to both send and receive ultrasound signals, like a combined microphone and speaker. Piezoelectric elements organized in an array, is referred to as an ultrasound \textit{probe}, and a special probe is typically designed for each medical imaging modality. An example of a probe for medical ultrasound imaging is shown in Fig.\,\ref{fig:us_grid}. %For cardiac ultrasound imaging the probe must not only be designed for a certain frequency range, its footprint also needs to be smaller than the intercostal window in order to avoid shadowing by ribs.

If the width of a piezoelectric element is less than the wavelength of the transmitted signal the element is said to be \textit{omnidirectional}, hence, energy will be distributed or sensed equally in all directions. When multiple omnidirectional elements are arranged in an array the transmitted pulse will get shaped into a beam\footnote{A beam refers to the pulse in space integrated across time.} of ultrasound by diffraction as the size (the \textit{aperture}) of the array grows. The angular width of this beam, and also the angular resolution of the array, is approximately given by \cite{AngelUltrasound}:
\begin{align}\label{eq:res}
\delta\theta_a = \frac{\lambda}{D} = \frac{c}{fD},
\end{align}
\nomi{$\delta\theta_a$}{Angular resolution of an aperture}where \nom{$\lambda$}{Wavelength} and \nom{$f$}{Frequency} is the ultrasound wavelength and frequency respectively, $c$ is the speed of sound in human tissue, and \nom{$D$}{Aperture size} is the aperture width. With an average speed-of-sound in human tissue of $1540\frac{\text{m}}{\text{s}}$, the angular resolution or opening angle of a 2 cm cardiac probe operating at 3 MHz is around 1.5\degree. The axial resolution is inverse proportional to the pulse length, and is typically better than the lateral resolution. The geometry of cardiac ultrasound imaging is depicted in Fig.\,\ref{fig:us_grid}. A line in an ultrasound image is generated by firing a ultrasound pulse or beam into the body and sample the received echo. We will refer to this amplitude trace as an \textit{image line}.

\subsection{Beamforming}
If time delays are applied on the signal emitted by the elements it is possible to steer and focus the ultrasound beam in space (Fig.\,\ref{fig:das}). This technique is known as \textit{beamforming}, and for ultrasound imaging it is applied both when transmitting and receiving ultrasound pulses. A cardiac probe is a phased probe, which means that the image is constructed from image lines which are gradually steered from $-\theta/2$ to $\theta/2$ radians. As depicted in Fig.\,\ref{fig:us_grid}, this results in an imaging sector which is $\theta$ radians wide. This is known as a 2D B-mode image (B for brightness). If image lines are acquired for multiple B-mode images which are tilted in the $yz$-plane we get what is known as a volumetric B-mode or a 3D ultrasound image. A probe for 3D imaging has several thousand elements arranged as a matrix in order to facilitate steering in both the $xz$ and $yz$-plane. In this thesis we will focus on 2D imaging when the topic is adaptive beamforming, and 3D imaging when the topic is adaptive visualization.

@@

The standard form of beamforming used on receive in ultrasound imaging is known as \textit{delay-and-sum} (DAS) beamforming: 
\begin{align}\label{eq:das}
z[n] = \sum_{m = 0}^{M-1}w_m^*x_m[n - \Delta_m[n]] = \vec{w}^H\vec{x}[n].
\end{align}
Here, \nom{$M$}{Number of array elements} is the number of elements in the array, $x_m$ is the per-element data, \nom{$\Delta_m$}{Per-element focusing and steering delays} is the per-element focusing and steering delay, \nom{$\vec{w}$}{Array weight vector (apodization)} is the array \textit{weight vector} or \textit{apodization}, $\vec{x}$ is a pre-delayed vector of element data, and $z[n]$ is the output image line in the direction given by $\Delta_m[n]$.  From now on we will omit the dependency on $\Delta_m[n]$ and $x_m$ will refer to delayed element data. A schematic overview of delay-and-sum array beamforming is depicted in Fig.\,\ref{fig:das}. If $p_o$ is the phase center of our array and $p$ is the point to which we want to steer and focus the ultrasound beam, we have to use the following delays:
\begin{align}
\Delta_m[n] =\frac{|p_o+p| - |p_m+p|}{c},
\end{align}
where \nom{$c$}{Speed of sound} is the speed of sound in human tissue, and $p_m$ is the position of the m'th element. On transmit the inverse set of delays has to be used. Also note that on transmit we have to decide where to steer and focus prior to transmitting a pulse. Transmit delays therefore do not vary within an image line.

\subsection{Sampling and processing complexity}
When generating a B-mode image we have to insure that the spatial and temporal sampling are sufficient. We have already seen that the lateral resolution of an ultrasound probe is given by (\ref{eq:res}).  Hergum \textit{et al.} \cite{Hergum2009} have further shown that the lateral frequency support for a system which both emit and receive energy is governed by the sum of the transmit ($D_{tx}$\nomi{tx}{Transmit}) and receive ($D_{rx}$\nomi{rx}{Receive}) aperture. The lateral resolution of what is known as an active imaging system is found to be
\begin{align}
\delta\theta = \frac{\lambda}{D_{tx} + D_{rx}}.
\end{align}
\nomi{$\delta\theta$}{Angular resolution of a two-way imaging system}\nomi{$D_{tx}$}{Aperture size on transmit}\nomi{$D_{rx}$}{Aperture size on receive}Then, in order to properly sample an imaging sector of $\theta$ radians we have to acquire 
\begin{align}
N_\theta = \frac{\theta}{\delta\theta}
\end{align}
\nomi{$N_\theta$}{Number of lines required for proper sampling of an imaging sector}image lines distributed across the imaging sector. If \nom{$r$}{Imaging range.} is the maximum range in our image, as shown in Fig.\,\ref{fig:us_grid}, we can acquire
\begin{align}
f_{\text{FPS}} = \frac{c}{2rN_\theta}
\end{align} 
images or frames per second (\nom{FPS}{Frames per second}). With our 2 cm cardiac probe we get a frame rate of around $55$ FPS when imaging a 70\degree{} sector with $r=15$ cm.
